### 2017-12-11 Writing/disparity_review.tex
review reframed

%Packages
\usepackage{pdflscape}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{dcolumn}
\newcolumntype{d}[1]{D{.}{.}{#1}}

@@

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$, Natalie Cooper$^2$, Philip Donoghue$^3$}\\
\noindent {\small \it 
$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\}
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

@@

%§1: disparity is observed everytwhere
Disparity is an old concept blabalbal coming from the observation that some groups of species are more similar/dissimilar than other.

It can be also applied between groups as a concept of ecological niches: a group occupying many niches will have a stronger disparity than a groups with less niches.

It is decoupled from diversity.

\bigskip
%§2: "phylogeny" of disparity

In palaeobiology, morphological disparity (the (dis)similarities between the morphologies of groups of organisms) concept as a tool for macroevolutionary studies stems to several seminal papers from the 90s: \cite{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution}.

This concept was then used in geometric morphometric and expanded to macroevolutionary studies in general (Harmon's dtt).

\cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.

\bigskip

%§3: a very popular concept

Disparity is really popular (Fig. \ref{Fig:GoogleOccurences}):

@@

\section{Why do we need disparity}
Which macroevolutionary questions are tackled through disparity analysis?

What are the fundamental questions we should use to answer disparity questions?

What are the questions we should not answer with disparity?

Disparity can be 
 -From a semantic view: morphospace or aspect of the morphospace?
 -From a biological view: what does an increase or a decrease in disparity represent?

But in the end, what \textit{is} disparity

\section{But what \textit{is} disparity?}

@@

To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in the following categories:

[DESCRIBE THE CATEGORIES HERE (from \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd})]

@@

\subsection{Disparity data}
Describing the many data inputs (from collected data to morphospace):

\subsection{Disparity metrics}
Describing the many metrics (what is actually measured):

@@

@@

\begin{itemize}
    \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data for the full organisms to build a PCO from the organism's pairwise distance as a morphospace. Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g. the sum of variance or/and the sum of ranges). Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.

@@

    \item \textbf{The ``\texttt{dtt}'' approach:} continuous data for the full organism, not ordinated, with a pairwise distance based disparity metric and phylogenetic correction.

@@

\section{What is missing}

How to compare disparity between groups?

@@

Hot to compare disparity between disciplines?

%TG: we can find solutions from other fields
It bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.

@@

\section{Acknowledgments}
The Royal Society.
I acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.

>>>>

%Packages
\usepackage{xcolor}
\usepackage{color,soul}
\usepackage{pdflscape}
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{float}
\usepackage{latexsym}
\usepackage{url}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
\usepackage[version=3]{mhchem}
\usepackage{ifthen}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{amsthm}
\usepackage{amstext}
\usepackage{enumerate}
\usepackage[osf]{mathpazo}
\usepackage{dcolumn}
\usepackage{lineno}
\usepackage{dcolumn}
\usepackage{hyphenat}
\usepackage[T1]{fontenc}
\usepackage{textcomp}
\newcolumntype{d}[1]{D{.}{.}{#1}}

@@

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^*$, Natalie Cooper$^2$, ..., Philip Donoghue$^3$}\\
\noindent {\small \it 
$^1$Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.\\}
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{guillert@tcd.ie}\\  
\vspace{1in}

@@

\noindent \hl{\textit{1: disparity is observed everywhere}}
Biodiversity is not a smooth gradient of forms, functions, ecologies and behaviours but it is rather variable and discontinuous.
This is at the base of Biological Sciences and is linked to the concept of disparity with the observation that some groups of species are more similar or dissimilar than others.
This diversity in morphologies is in fact proper in nature and decoupled from taxonomic diversity: some groups may exhibit low diversity of species by high diversity of shapes \citep[or the other way around]{ruta2013,hopkinsdecoupling2013}.

\noindent \hl{\textit{2: historical evolution of disparity}}
Morphological diversity (hereafter disparity) is based on the observed the (dis)similarities between the morphologies of groups of organisms.
In palaeobiology, this concept, as a tool for macroevolutionary studies, stems from several seminal papers from the 90s describing and questioning how the diversity of life forms arose and evolved \citep{gould1989wonderful,gould1991disparity,briggs1992morphological,Wills1994,Foote01071994,Foote29111996,jernvall1996molar,foote1997evolution}.
In parallel, disparity analysis have been extended with modifications to macroevolutionary analysis based on continuous traits or geometric morphometric.
The former usage of data was popularised following \cite{Harmon961} and its implementation in \texttt{geiger} \citep{geiger2008}; while the latter was popularised following \citep{zelditch2012geometric} %Probably older
and implementations in \texttt{geomorph} (\citealt{adams2013geomorph, adams2017geometric} and precursory work from \citealt{claude2008morphometrics}).

\noindent \hl{\textit{3: the great popularity leads it to be used in many methods}}

\cite{prentice2011} define disparity as: ``a term widely (albeit not always consistently) used to describe the range of forms in a group of organisms, or the difference among different body plans''.
Disparity is really popular (Fig. \ref{Fig:GoogleOccurences}):

@@

\section{Why do we need disparity}

\noindent \hl{\textit{These points should be developed during the meeting}}

Which macroevolutionary questions are tackled through disparity analysis?
\begin{itemize}
    \item Bodyplan evolution through time
    \item Why does some morphologies exist and other not?
    \item tempo and mode of innovation of genuine novelty
    \item evolutionary competition
    \item effects of mass extinctions
    \item ecological niches
\end{itemize}

To answer these question, disparity has been used as a proxy for the biological increase/decrease of distinguishable morphological variation.
These proximal multidimensional spaces have been described as the morphospace, the eco-space, the functional space, the cladisto space, the developemental space, the allometric space, etc \citep[see][and references therein]{Hopkins2017}.
This multitude of multidimensional spaces arising from the observation of variable and discontinuous morphology in Life arises the question of their equivalence and their pertinence.
Are the description of these spaces all equivalent?
If disparity is the method or the metric describing these spaces then what \textit{is} disparity?

\section{What \textit{is} disparity (really)?}

@@

To reduce the amount of categories for the 230 recorded methods, we concatenated different methods in a smaller number of categories (see supplementary materials and \url{https://github.com/TGuillerme/Disparity_Working_Group/blob/master/Analysis/data_cleaning.Rmd}).

@@

\hl{[ADD Biological group, full organism and phylogeny in the supplementary]}

\noindent \hl{\textit{These points should be developed during the meeting}}

\subsection{Disparity data}
The data used for disparity methods comes from three main sources:
\begin{itemize}
    \item Continuous data such as limb or body measurements \citep{slaterCetacean}.
    \item Discrete morphological characters \citep[sometimes referred to as ``Cladistic'' characters; ][]{Brusatte12092008}.
    \item Geometric morphometric data which is generally the Procrustes transformation of 2D or 3D landmarks \citep{cooney2017mega}.
\end{itemize}

However, data was by far not limited to these three categories (e.g. colours \citealt{maia2013key}; metabolic rate \citealt{nespolo2016studying} or chemicals signal \citealt{garcia2017heterogeneous}).

\subsection{Morphospaces}
\begin{itemize}
    \item PCAs [CITE].
    \item PCOs [CITE]
    \item Pairwise distance matrices \citep[no ordination; ][]{Harmon961}.
\end{itemize}

But also RWA [CITE] or Raup-spaces [CITE].

\subsection{Disparity metrics}
Throughout the 230 analysed papers, we found 103 unique combinations of metrics!

\begin{itemize}
    \item Distances measurements between species or from a certain point in the morphospace [CITE].
    \item Ranges and variances of each axis of the morphospace \cite{Wills2001,Ciampaglio2001}
    \item Distance between species based on the pairwise distance matrix (not the ordinations) [CITE].
\end{itemize}

But also metrics based on the volume [CITE], the characters dissimilarity [CITE] or the coordinates of some axis of the ordination [CITE]

@@

\begin{itemize}
    \item Variance based (ANOVA, etc.) [CITE].
    \item Correlation based [CITE].
    \item Regression based (PGLS, etc.) [CITE].
\end{itemize}

But also discriminant analysis [CITE] and clustering [CITE].

@@

\begin{itemize}
    \item \textbf{The ``\texttt{Claddis}'' approach:} this group of methods uses discrete morphological data (sometimes referred to as ``Cladistic'' characters) for the full organisms to build a PCO from the organism's pairwise distance as a morphospace. Disparity is then often measured as a variation of the ordinated matrix dimensions' variances or ranges (e.g. the sum of variance or/and the sum of ranges). Hypothesis are often tested using multivariate ANOVAs on the pairwise distance matrix or by simply comparing the confidence intervals overlap of the disparity from different groups.

@@

    \item \textbf{The ``\texttt{dtt}'' approach:} this method can directly use continuous or discrete data for the full organism without any ordination (but not necessarily), and will measure disparity as the average pairwise distance between species (whether euclidean or any other type of distance). Hypothesis of higher/lower disparity can then be measured using null evolutionary models.

@@

\section{Expanding disparity}

How to compare disparity between groups? Is disparity relative?

@@

Can we really say things about competition when looking at disparity in a single group?

Are all characters equal? Character contingency suggests otherwise

What is the relationship between disparity and tree shape?

Do we have a null model for investigating disparity?

Can we really say things about competition when looking at disparity in a single group?

\noindent \hl{\textit{Disparity in other fields}}

In ecology, disparity bears strong parallels with $\beta$-diversity in ecology (a measure of ecological communities (dis)similarity): one biological observation described by a vast array of metrics \citep{baselga2010partitioning, anderson2011navigating, donohue2016navigating}.

@@

\section{Acknowledgments}
The Royal Society.
TG acknowledge support from European Research Council under the European Union's Seventh Framework Programme (FP/2007 – 2013)/ERC Grant Agreement number 311092 awarded to Martin D. Brazeau.
